Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-07-28 sports_analytics.Rmd
Some greg football changes.

A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.
The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes. 
This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.

@@

@@

Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.

(A paragraph on applications of expected points on decision making, fourth down strategy, with an example)

4th down strategy papers based on expected points.  
@lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.  
<!-- @Yam2019: Uses Causal inference techniques to examine the effects of 4th down decision making in the NFL.   -->
<!-- @Romer2006: Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game.   -->
@pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.  
<!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->

@Alamar2010 Evaluation of play calling in the NFL based on expected points.  
https://www.degruyter.com/document/doi/10.2202/1559-0410.1235/html

@urschel2011 Examines decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.  Shows that teams are moderately risk and loss averse.

@White2002 Evaluated quaraterback usein gtiered multinomial regression in an expected points framework. They use a simple expected points model with down, to go, and to goal as predictors.   
https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf

@@

In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football. 
These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.

@@

On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game. 
This model assumes independence of observations and constant variability in win probability.

>>>>

A more in-depth overview of the history of expected points in sports is provided in @Yurko2019nflwar (Section 1.1). 
Most importantly, this paper uses publicly available data provided by the **nflscrapR** package [@R-nflscrapR] to model the expected points on a play-by-play level in football.

The authors introduce a multinomial logistic regression approach, which takes into account the current down, time remaining, yards from endzone, yards to go, and the indicators for goal to go and whether the time remaining in the half is less than two minutes. 
This model estimates the probabilities of the following possible scoring outcomes after each play: no score, safety, field goal, and touchdown for both the offensive and defensive teams, all of which have a point value.

@@

@pelechrinis2019 develops an expected points framework in the same spirit as previous work, but accounts for the strength of the opponents in their method.  They state that by failing to account for opponent strength appropriately, about 124.8 points per team each season (or about 3.8 wins per season) are not credited correctly.

@@

Recently, @lopez2020bigger presented an introduction to NFL tracking data, and as an example they examine 4th down behavior.  Rather than looking at the traditional distance to go, which is rounded, they have the exact position of the ball leading to an exact distance to go.  They demonstrate than teams are still conservative when choosing to go for it or punt on 4th down, but the effect may be overstated.

### Other applications of expected points in American football
@White2002 presents a tiered logistic regression method that can be, in general, applied to regression setting with a polychotomous response.  Using this technique, they present an example where they estimate the value of NFL plays using a simple expected points model with down, to go, and to goal as predictors.  They then use these results to evaluate and rank NFL quarterbacks.

@Alamar2010 uses an expected points framework to evaluate play calling in the NFL.  However, rather than evaluate each play on its own, the framework presented here evaluates the play in the context of the drive.  Based on play-by-play data from 2005 through 2008, they find that teams are underutilizing passing plays in some situations.

<!-- (A paragraph on applications of expected points on decision making, fourth down strategy, with an example) -->

<!-- 4th down strategy papers based on expected points.   -->
<!-- @lopez2020bigger: Talks about how tracking data is important for a more nuanced discussion of 4th down decisions.   -->
<!-- @Yam2019: Uses Causal inference techniques to examine the effects of 4th down decision making in the NFL.   -->
<!-- @Romer2006: Examines 4th down decisions in the NFL using expected points.  Focuses only on the first quarter.  Teams don't go for it enough is they are trying to maximize their probability of winning the game.   -->
<!-- @pelechrinis2019 Present an expected points framework where they adjust for the strength of the defense.   -->
<!-- New york times article: https://www.nytimes.com/2014/09/05/upshot/4th-down-when-to-go-for-it-and-why.html?_r=0 -->

<!-- @Alamar2010 Evaluation of play calling in the NFL based on expected points.   -->
<!-- https://www.degruyter.com/document/doi/10.2202/1559-0410.1235/html -->

@urschel2011 Examines decsions on kick offs (surprise on sides or regular kick off AND accept touchback or return kick) using an expected points frame work.  Shows that teams are moderately risk and loss averse.
https://www.eng.buffalo.edu/~jzhuang/Papers/JQAS_NFL_2011.pdf

@@

In recent years, a number of statistical methods have been used to build well-calibrated win probability models in American football.

These are flexible models that have high predictability, can account for nonlinear interactions between the explanatory variables, require few assumptions, and produce feature importance scores.

@@

On a related note, @Deshpande2016 extend @Stern1994's framework by applying it in a Bayesian setting. 
In this paper, the authors propose a Bayesian linear regression model to assess the impact of individual players on their team's chance of winning at any given time of a basketball game.

This model assumes independence of observations and constant variability in win probability.
